# Data transformation

## define function
```{r}
load("./data/RData/data_MMASH.RData")
library(chron)
# create function to transform string object to time object for further calculation
string2time <- function(x, format = "%H:%M"){
  return(strptime(x, format = format)) # times(sub(".*\\s+", "",  
}
```

## Actigraph
Actigraph.csv - accelerometer and inclinometer data recorded throughout the day:
Axis1: Raw Acceleration data of the X-axis expressed in Newton-meter.
Axis2: Raw Acceleration data of the Y-axis expressed in Newton-meter.
Axis3: Raw Acceleration data of the Z-axis expressed in Newton-meter.
Steps: number of steps per second.
HR: beats per minutes (bpm).
Inclinometer Off: values equal to 1 refer to no activation of the inclinometer. The values are reported per second.
Inclinometer Standing: values equal to 1 refer to the standing position of the user, while 0 refers to other user positions. Values are reported per second.
Inclinometer Sitting: values equal to 1 refer to the sitting position of the user, while 0 refers to other user positions. Values are reported per second.
Inclinometer Lying: values equal to 1 refer to the lying position of the user, while 0 refers to other user positions. Values are reported per second.
Vector Magnitude: vector movement derived from raw acceleration data expressed in Newton-meter.
day: 1 and 2 refer to the first and second day of data recording, respectively.
time: day time when the heartbeat happened (hours:minutes:seconds)

```{r}


```

## Activity
calculate duration for each activity
```{r}

activity$Start = string2time(activity$Start)
activity$End = string2time(activity$End)
activity$Duration = difftime(activity$End,activity$Start)

```