# Missing values

Missing information includes - 

(1) file information missing: user_11 is missing sleep.csv information

(2) No clock genes and hormones concentrations data was provided for User_21 due to problem in the salivary samples that do not permit to analyse it.

(3) missing age for user_18 in user_info

(4) user_13 is missing questionnaire

Since the aim of our project is to define a metric for sleep quality and identify variables with that affect it, user 11 was excluded from all the analysis performed. 

## Missing Value Plot

Here we revise the function we created in problem set 4 to plot the missing patterns. The number of missing values are very few (mentioned above). We generated missing value plots only for `questionnaire` and `activity` as they seemed to exhibit some pattern.
As shown in the plots, only one row with missing values exist in `activity` and `questionnaire`, respectively.
```{r}
# missing value plot function from problem set 4
library(dplyr)
library(ggplot2)
library(tidyverse)
library(patchwork)
library(ggnewscale)

missingValuePlot <- function(df, mode = "count", title = 'Missing value patterns' ){
  # create the missing value dataframe, depends on the mode chosen
  if(mode == 'count'){
    # construct dataframe/table base on counts
    missing_patterns <- data.frame(is.na(df)) %>%
    group_by_all() %>%
    count(name = "count", sort = TRUE) %>%
    ungroup()

  }else if(mode == 'percent'){
    # construct dataframe/table base on percentages
    missing_patterns <- data.frame(is.na(df)) %>%
    group_by_all() %>%
    count(name = "count", sort = TRUE) %>% 
    ungroup()
    missing_patterns$count <- missing_patterns$count/sum(missing_patterns$count) * 100
    
  }else{
    warning('Please select valid options: "mode" or "percent"')
  }
  
  
  # plot here
  
  missing_count = missing_patterns$count
  missing_patterns$count = NULL
  
  # side plot for column
  col_count = data.frame(count = colSums(missing_patterns * missing_count), col = names(missing_patterns))
  col_count = arrange(col_count, -count)
  
  missing_patterns <- data.frame(lapply(missing_patterns, as.numeric))
  # create vector to highlight complete cases
  complete_case = c()
  for (pat in c(1:dim(missing_patterns)[1])){
    if (all(missing_patterns[pat,] == 0)){
      complete_case = c(complete_case, 1)
    }else{
      complete_case = c(complete_case, 0)
    }
  }
  complete_case = data.frame(complete = complete_case, pattern = c(1:dim(missing_patterns)[1]))
  
  # side plot for row
  row_count = data.frame(count = missing_count, pattern = complete_case$pattern)
  row_count <- dplyr::left_join(row_count, complete_case, by = 'pattern')
  
  # plot side bar plot for column
  if(mode == 'count'){
    p_col = col_count %>% ggplot(aes(x = reorder(col,-count), y = count)) + 
      geom_col(fill = 'blue', alpha = 0.7, show.legend = FALSE) + 
      scale_x_discrete(labels = abbreviate) +
      theme_bw() + 
      theme(panel.grid.minor.x = element_blank(),
            panel.grid.major.x = element_blank(), 
            axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
      labs(title = title, x = " ", y = "num rows missing")
  }else{ # percentage
    p_col = col_count %>% ggplot(aes(x = reorder(col,-count), y = count)) + 
      geom_col(fill = 'blue', alpha = 0.7, show.legend = FALSE) + 
      scale_x_discrete(labels = abbreviate) +
      ylim(0, 100) + 
      theme_bw() + 
      theme(panel.grid.minor.x = element_blank(),
            panel.grid.major.x = element_blank()) +
      labs(title = title, x = " ", y = "% rows missing")
  }
  
  # plot side bar plot for row
  if(mode == 'count'){
    p_row = row_count %>% ggplot(aes(x = factor(pattern), y = count, alpha = factor(complete))) + 
      geom_col(fill = 'blue', show.legend = FALSE) + 
      scale_x_discrete(limits = rev) +
      coord_flip() + 
      theme_bw() +
      scale_alpha_manual(values = c(0.7,1)) +
      theme(panel.grid.minor.y = element_blank(),
            panel.grid.major.y = element_blank()) +
      labs(x = ' ', y = 'row count')
  }else{ # percentage
    p_row = row_count %>% ggplot(aes(x = factor(pattern), y = count, alpha = factor(complete))) + 
      geom_col(fill = 'blue', show.legend = FALSE) + 
      scale_x_discrete(limits = rev) +
      ylim(0, 100) +
      coord_flip() + 
      theme_bw() +
      scale_alpha_manual(values = c(0.7,1)) +
      theme(panel.grid.minor.y = element_blank(),
            panel.grid.major.y = element_blank()) +
      labs(x = ' ', y = '% rows')
  }
  
  # main plot
  # temp = missing_patterns
  missing_patterns$pattern = c(1:dim(missing_patterns)[1])
  df = missing_patterns %>% pivot_longer(!pattern, names_to = "column", values_to = "missing") %>% dplyr::left_join(complete_case)
  
  p_main = df %>% 
    ggplot(aes(x = factor(column,levels = rownames(col_count)), y = factor(pattern,levels = rev(rownames(row_count))), fill = factor(missing), alpha = factor(complete))) + 
    geom_tile(color = 'white', lwd = 0.4, linetype = 1, show.legend = FALSE)+  
    scale_x_discrete(labels = abbreviate) +
    scale_fill_manual(values = c('grey', 'purple')) +
    scale_alpha_manual(values = c(0.5,1)) +
    theme(panel.background = element_blank(), axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))+
    ylab('missing pattern') +
    xlab('variable')
  
  # patchwork
  layout <- "
  AA#
  BBC
  BBC
  "
  # final plot
  p_col + p_main + p_row + plot_layout(design = layout)
    
}
```

There is only one missing value in "End" variable which is defined as the time of the day when the event or activity concluded. For example, if a user smoked from 10:30 AM to 10:40 AM then `End` would be set to 10:40 AM. Except this one missing value, all others are complete cases.

```{r}
# load data from previous stage
load("./data/RData/data_MMASH_cleaned.RData")
# plot
# missingValuePlot(actigraph, title = 'Missing value patterns for actigraph')
missingValuePlot(activity, title = 'Missing value patterns for activity')

# missingValuePlot(RR, title = 'Missing value patterns for RR')
# missingValuePlot(saliva, title = 'Missing value patterns for saliva')
# missingValuePlot(sleep, title = 'Missing value patterns for sleep')
# missingValuePlot(user_info, title = 'Missing value patterns for user_info')
```


We generated this plot after removing user 13 so as to not distort the actual missing value patterns for users that have a record for `questionnaire`. Similar to `activity`, this file category has only two patterns - complete cases and two missing values that correspond to the levels of a variable "PANAS". PANAS is defined as the positive and negative affect schedule. It gives a score rating between 5 and 50 for both positive and negative emotions and higher the value, the higher is the perceived emotion. There are only one rows that are not complete.

```{r}
missingValuePlot(questionnaire, title = 'Missing value patterns for questionnaire')
```