# Results
The results chapter is broadly divided into two sections - one for each of the goals of this project. The first section will document all preliminary findings related to the sleep measurements and the procedure of defining a metric for quantifying quality of sleep.

## Metric Formulation
### Preliminary Observations
Before we delved into the data analysis part we wanted to see how each of the variables in the sleep.csv file is distributed. Upon digging into the data we realized that user 11 had no measurements for his sleep. The authors of the data did not mention a reason as to why there were no records for that particular user. Since our problem statement is about quanitfying sleep quality and the factors affecting it, we thought it only logical to remove user 11 from our analysis altogether.

Formulating a metric requires an extensive analysis of the spread of the data and each variable in it. It is also important to determine correlated variables to avoid duplication and/or multicollinearity while defining the metric. Below is a heatmap we generated for variables that do not have the format of a time stamp (for example hh:mm). We made the following observations - 

(1) Latency does not seem to be correlated with anything which is interesting as it is defined as the time taken for the user to fall asleep after he gets into bed
(2) Interestingly enough, the total sleep time is not correlated with neither Number of Awakenings and Average Awakening Length. One would suspect that the form is negatively correlated to the latter two but that is not the case as these correlation coefficients are statistically significant with a default critical value of 0.05
(3) Movement, Fragmentation and Sleep Fragmentation index are highly correlated and this relationship is statistically significant. This observation is not surprising as sleep fragmentation index is defined as the ratio of movement index and fragmentation index
(4) Wake After Sleep Onset (WASO) is strongly correlated to almost every variable with the direction being positive except for Latent Efficiency


```{r}
library(Hmisc)
library(corrplot)
library(gplots)
load("./data/RData/data_MMASH_cleaned.RData")
sleep_data <- read.csv('data/raw/multilevel-monitoring-of-activity-and-sleep-in-healthy-people-1.0.0/main_sleep_transformed.csv')
# adding user id
sleep$user.id <- as.numeric(unlist(map(strsplit(sleep$user, split = "_"), 2)))
sleep <- sleep[order(sleep$user.id),]

sleep_data <- as.data.frame(t(sleep_data))
colnames(sleep_data) <- sleep_data[1, ]
sleep_data <- sleep_data[-1, ]

#removing variables with time entries of format hh::mm
sleep_subset <- subset(sleep_data, select = -c(1,3,5))

# converting all values that are not time readings to numeric type
sleep_subset[,4:length(sleep_subset)] <- sapply( sleep_subset[,4:length(sleep_subset)], as.numeric)
mydata <- sleep_subset[,4:length(sleep_subset)]

mydata.rcorr = rcorr(as.matrix(mydata))
mydata.coeff = mydata.rcorr$r
mydata.p = mydata.rcorr$P
palette = colorRampPalette(c("darkblue", "white", "darkred")) (20)
heatmap.2(x = mydata.coeff, symm = TRUE, labRow = abbreviate(rownames(mydata.coeff), minlength = 24), labCol = abbreviate(colnames(mydata.coeff), minlength = 24), col = palette, margins=c(10,16), srtCol=45,main = "heatmap", density.info="none", trace="none")

mydata['user'] <- c(1,2,3,4,5,6,7,8,9,10,12,13,14,15,16,17,18,19,20,21,22)
```

### Distribution of Variables
The clustering of heatmap showed that there are 3 clear clusters which are consistent with the observations we made. To avoid multicollinearity, we decided to pick 1-2 variables from each of the clusters depending on the strength and statistical significance of the correlation coefficient. Let's denote our metric with M and is defined as a characterization of quality of sleep. Intuitively, M should be higher for users that are able to sleep through the night and do not have any trouble falling asleep. Expanding this thought process further and combining it with the correlation analysis, we came up with the following associations - 

(1) M $\alpha$ Total Sleep Time (TST)
(2) M $\alpha$ $\frac{1}{SleepFragmentation Index}$
(3) M $\alpha$ $\frac{1}{Latency}$
(4) M $\alpha$ $\frac{1}{AverageTimeSpentAwake}$

Note that (2) is a ratio while others have units "hh:mm". So, keeping that in mind we combined all the above relationship as - 

\begin{align*}
M \alpha (\frac{TST}{AverageTimeSpentAwake} + \frac{1}{SleepFragmentationIndex}).\frac{1}{Latency}
\end{align*}

The units of the right hand side are "per hh:mm" (because of addition of latency). Since this expression was constructed using just the correlation, to establish a stronger relationship i.e. equality, it is important to look at the distribution of each of these variables. It might happen that even though intuitively one might think that a variable should have an effect on the sleep quality but if in reality the distribution of that variable turns out to be narrow then including it would not be a good idea. We would like to stress that the number of samples available to us are very less to do this sort of analysis without any selection bias i.e. inferences might hold for this small cohort but not necessarily for a bigger cohort.

We defined sleep ratio to be the ratio of TST and Average Time Spent Awake. The latter was calculated by simply multiplying Number of Awakengings with the Average Awakening Length. The box plot below shows the distribution of the Sleep Ratio and Sleep Fragmentation Index. Indeed both ratios seem to have a positive variance. 
```{r}
library(dplyr)
mydata['Average Time Awake'] = mydata$`Number of Awakenings`*mydata$`Average Awakening Length`
mydata['Sleep Ratio'] = mydata$`Total Sleep Time (TST)`/mydata$`Average Time Awake`
df <- dplyr::select(mydata,c('Sleep Ratio','Sleep Fragmentation Index'))

ggplot(stack(df), aes(x = ind, y = values)) + geom_boxplot() +labs(x = "", y = "Ratio Values", title = "Distribution of Sleep Ratio and Sleep Fragmentation Index in Cohort") 
```

Since Latency is the time taken by the user to fall asleep once he gets into the bed, we wanted to visualize it separately. We noticed that a lot of the users had a latency time of 0 mins. This seems a little absurd given its definition so we assumed that any user that fell asleep within a minute had a latency of 0. This creates a problem since the metric is inversely proportional to latency and if it has a value of zero, it would mean the absolute value of M would be infinity. Clearly, this isn't possible and so when replacing proportionality with equality, we add a dummy term to latency. 


```{r}
ggplot(df, aes(mydata$user, mydata$Latency)) + geom_point() + scale_x_continuous(breaks = scales::pretty_breaks(n = 17)) + labs(x = "User ID", y = "Time in Mintues", title = "Latency per User") + theme_bw()
```

We define the sleep quality of a user as - 

\begin{align*}
M = (\frac{TST}{AverageTimeSpentAwake} + \frac{1}{SleepFragmentationIndex}).(\frac{1}{Latency + 1})
\end{align*}

The dummy term takes care of the case when Latency is 0. 

### Distribution of Sleep Quality Metric

We calculated the sleep quality metric for each of the users and visualized the results as a boxplot to be able to discern its distribution. The lowest value is 0.5902 corresponding to user 12 and the largest is 16.33 corresponding to user 12, who is also an outlier. We are particularly interested in investigating users that fall below the lower hinge and above the upper hinge. The former group includes users 2, 3, 8, 10 and 16. While the group with M > 6.6874 includes users 1, 12, 14, 20 and 21. The rest lie within the 25th and 75th percentile of the metric data.

```{r}
library(tidyverse)
# calculating the metric
mydata['M'] = ((mydata$`Total Sleep Time (TST)`/mydata$`Average Time Awake`)+(1/mydata$`Sleep Fragmentation Index`))*(1/(mydata$Latency + 1))

#adding M to sleep data
sleep$M <- c(rep(mydata$M[1],2), mydata$M[2:length(mydata$M)])

# ggplot(mydata, aes(`user`, `M`)) + geom_point() + scale_x_continuous(breaks = scales::pretty_breaks(n = 17)) + labs(x = "User ID", y = "Units (per minute)", title = "Sleep Quality of each User") + theme_bw()

ggplot(stack(dplyr::select(mydata, c('M'))), aes(x = ind, y = values))  + geom_boxplot() + geom_jitter() +labs(x = "Sleep Quality", y = "Units (per minute)", title = "Distribution of Sleep Quality") 

fivenumnames <- c("min", "lower-hinge", "median", "upper-hinge", "max")
l <- mydata$M
fivenum(l) %>% set_names(fivenumnames)
```

Having defined the metric, we wanted to do a sanity check that it indeed is capturing the quality of sleep. The only way to do that was to compare the values of Efficiency variable to M. Efficiency is defined as the percentage of Total Sleep Time upon total sleep in bed. The reader might think that the metric defined above could have been made way simpler if Efficiency was included but we had motivation to not include it altogether. One of the things that pops out is the Efficiency is not strongly correlated to both of the variables TST and Total Minutes in Bed. This seems absurd and since we couldn't explain why this was happening we ultimately decided to remove efficiency from the equation and instead use it as a sanity check. It is only logical to think Efficiency and sleep quality should be positively correlated.  

To test the relationship, we generated a log-log scale plot. The motivation behind using a log10 scale over the normal scale was first to normalize the values and second to see how the magnitudes of the two variables are related. The regression model is statistically significant with a p-value of 3.681e-05 and the positive slope indicates that as the sleep quality increases, so does the efficiency. However, caution needs to be exercised while making this conclusion as the adjusted R square value is 0.6 which means that the model captures less than 70% of the variance. 

In summary, the metric defined by the formula above seems to be capturing the information that directly affects the sleep quality of a user. Moreover, the statistical significance combined with the positive slope of the regression model and line prove to be a sanity check for our metric.

```{r}
library(ggpubr)
ggplot(mydata, aes(x = log(`M`, 10), y = log(`Efficiency`, 10))) + geom_point() + geom_smooth(method="lm") + labs(title = "Efficiency vs Sleep Quality (log10 scale)") + stat_regline_equation(label.x=-0.3, label.y=1.99) + stat_cor(aes(label=..rr.label..), label.x=-0.3, label.y=1.98)

model <- lm(log(mydata$Efficiency, 10)~log(mydata$M, 10), header = T)
summary(model)
```


## Establishing Direct/Indirect Effects on Sleep Quality
Having established a metric and quantified its distribution, we want to identify variables that have a direct or indirect relationship with sleep quality. These relationships might not necessarily be causal in nature as correlation does not imply causation. 

This section is divided further into subsections for each of the 7 files in the dataset. Since some of the data is redundant, we decided to analyze only 5 out of the 7 - sleep records, user information, actigraph measurements, activities performed and the saliva samples.

### In/Out Bed Time is Related to Sleep Metric?


```{r}
sleep %>% slice(-c(1)) %>% ggplot(aes(`In.Bed.Time`,`M`)) + geom_point() + labs(title = "Sleep Quality vs In Bed Time", x = "Time", y = "Sleep Quality (M)") + theme_bw()

sleep %>% slice(-c(1)) %>% ggplot(aes(`Out.Bed.Time`,`M`)) + geom_point() + labs(title = "Sleep Quality vs Out Bed Time", x = "Time", y = "Sleep Quality (M)") + theme_bw()

```